---
title: "Project Report Outline"
format: docx
editor: visual
---

## Executive summary

Project overview. Written by all participants.

## Introduction

-   Background on invasive species
-   Background on NYS DEC invasive species programs and the PRISM system
-   Information and background on the NYS Invasive Species lists

## Region of study

-   Information about the Lower Hudson region (as defined by LHPRISM)
-   Discussion of the climate and ecology of the region

## Overview of the LHPRISM data set

-   Description of the data set and its contents
-   Description of data collection methods (will require communication with LHPRISM staff)
-   Summary statistics and figures about major components of the data set (e.g., number of records, number of species, etc.)
-   Should we put analyses of correlations and relationships among the variables in this section?

## Spatial distribution of the data

-   Maps of the various locations and some relevant summary information
    -   Number of records per location
    -   Number of species per location
    -   Some summary of coverage percentages?

## Analysis of time / treatment effort required to eradication

-   This will be done on a subset of species
    -   How did we choose these species?
        -   accelerated growth, previous exploratory analysis file
            -   HEMA17 - Giant Hogweed
            -   PUMO - Kudzu
            -   CYSC4 - Scotch Broom
            -   SAGL12 - Sticky Sage
    -   Brief overview of ecology of these particular species (add known methods of treatment, previous recommendation)
        -   HEMA17 - Giant Hogweed (Heracleum mantegazzianum)
            -   native to Eastern Europe/Western Asia, highly invasive in NE USA
            -   thrives in rich, moist soil, prefers sunlight and tolerant of partial shade
            -   found in river banks, streams, floodplains, roadsides + trafficked areas
            -   biennial lifecycle, plants 2+ yrs old 10-15ft tall
            -   1 plant can produce 20,000-50,000 seeds, spreading by water/wind and human activity
            -   forms dense monocultures, increases erosion risk due to dying back in winter, leaves soil exposed along riverbanks
            -   phototoxic sap - sever skin burns with sun exposure
            -   treatment recommendation: chemical treatment in early spring, mechanical removal for small infestations
        -   PUMO - Kudzu
            -   native to East Asia, widespread in SE USA and migrating N
            -   thrives on roadside, forest edges, warm and humid climates, prefers full sun, climbs to canopy
            -   perennial vine, grows up to 1ft/day in optimal conditions
            -   spreads vegetatively, rooting at nodes
            -   forms dense monocultures, causes tree death by shading and structural stress
            -   alters nutrient cycle by nitrogen fixation
            -   treatment complex, requires multi-modal approach over multiple years
        -   CYSC4 - Scotch Broom
            -   Native to Europe, widespread invasiveness in PNW, California, and East Coast USA
            -   prefers sandy soil and full sun, found on roadsides, logged forests, grassland
            -   perennial shrub, lives up to 10-15 yrs
            -   1 plant produces 1,000s of seeds
            -   dense thickets, reduces habitat quality for native species, alters nutrient cycle by nitrogen fixation
            -   seeds viable for 30+ yrs in soil makes eradication difficult
            -   hand pulling for small infestations, chemical spray (glyphosate, triclopyr, basal bark treatment, stem cutting, foliar spray
        -   SAGL12 - Sticky Sage
            -   native to Europe and West Asia
            -   prefers moist, well-drained soil, found in woodlands, shaded slopes, forest edges
            -   perennial herb, blooms late summer
            -   low ecological impact, coexists with native species
            -   known treatment: handpulling for young pop
-   Description of the methods used in this analysis
    -   change in total plant count over consecutive visits
    -   visit_date \* visit_status_num
    -   maybe see if there are any patterns when sorting by herbicide?
    -   how many times did they need to visit a plot before eradication: look at total number of visits, and population numbers
-   Description of the results of this analysis
    -   HEMA17

        ```{r}

        ```

    -   PUMO

        ```{r}

        ```

    -   CYSC4

        ```{r}

        ```

    -   SAGL12

        ```{r}

        ```

## Conclusions and recommendations
