---
title: "Project Report Outline"
format: html
editor: visual
bibliography: references.bib
---

## Executive summary

Project overview. Written by all participants.

```{r package_setup}
## Get packages
library(pacman)
pacman::p_load(here, tidyverse, janitor, lubridate, leaflet, readxl, knitr)

## Get the project level directory path
proj_path <- here()

## Define a color blind friendly palette
## These are from - http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r data_import}
## Read in the data file
## In this code chunk, we will read in the data file. You may have to edit this call, depending on where you have saved your version of the Excel file.

lhprism_data <- read_csv(paste0(proj_path, "/data/LHPRISM-Data.csv"), 
    col_types = cols(`iMap Treat` = col_number(), 
        GROSS_TREAT_SQFT = col_number())) %>%
  clean_names()

## Edit the order of levels for some of the variables
lhprism_data <- lhprism_data %>%
  mutate(native = factor(native, levels = c("low", "medium", "high")),
         non_nat = factor(non_nat, levels = c("low", "medium", "high")))

## Create a new species code column
lhprism_data$species_code <- str_extract(lhprism_data$site_code, "(?<=_)[^_]+(?=_)") 
## Remove plant column
lhprism_data$plant <- NULL

## Create a new site column using the species_code column
lhprism_data$site <- str_extract(lhprism_data$site_code, "^[^_]+")

## Remove all rows that have NA for site
lhprism_data <- lhprism_data %>%
  filter(!is.na(site))

# Make a column for number of visits for each Site Code.
lhprism_data <- lhprism_data %>%
  group_by(site_code) %>%
  mutate(tot_visits = length(unique(visit_date)))


```

## Introduction

Ecosystems around the world face many threats — things like habitat loss, climate change, and pollution [@gentili2021]. One problem that people often overlook is the spread of invasive species [@usda2024]. In New York State, this isn't just a future worry. It's something that scientists, state agencies, and local communities deal with every day [@nysdec2024].

This report uses field data from the Lower Hudson Partnership for Regional Invasive Species Management (LHPRISM) to look at where invasive species are found in the Lower Hudson Valley and how they are being managed [@nysdec2024]. To understand why this data matters, it helps to know the bigger picture: what invasive species are, how New York State organizes its response, and what official plans guide decisions about which species to focus on.

### Background on Invasive Species

At its most basic, an invasive species is any organism — plant, animal, insect, fungus, or microbe — that establishes itself outside its native range and causes meaningful harm to the environment, economy, or human health [@mack2000; @usda2024]. The word "invasive" is doing a lot of work in that definition. Not every non-native species causes problems; in fact, most introduced species fail to survive at all in new environments [@mack2000]. They arrive without the adaptations needed to handle unfamiliar climates, predators, and food sources, and they quietly die out. The ones that do establish themselves and thrive are the exception — but when they do, the consequences can be severe and long-lasting [@gentili2021].

What makes certain species so successful in new environments often comes down to a lack of natural checks [@mack2000]. Back in their home range, they are kept in balance by predators, competitors, parasites, and other ecological forces that have co-evolved alongside them over thousands of years. When they land somewhere new, those controls disappear [@mack2000]. Native species in the invaded area have no evolutionary history with this newcomer and may not recognize it as a threat, compete with it effectively, or survive in its presence [@usda2024]. The invasive species, freed from its usual constraints, can reproduce rapidly, consume resources aggressively, and spread far beyond where it first appeared [@gentili2021].

The pathways through which invasive species move around the world are closely tied to human activity [@mack2000]. Global trade is one of the biggest drivers — cargo ships, shipping containers, and pallets of imported goods routinely carry hitchhiking organisms across continents and oceans [@usda2024]. The pet trade has introduced countless exotic animals into new regions, and when those animals escape or are deliberately released, some find conditions favorable enough to establish wild populations [@usda2024]. Ornamental horticulture has a long history of introducing plants that eventually escape garden boundaries and spread into natural areas [@mack2000]. Even recreational activities like hiking, boating, and fishing can accidentally transport seeds, larvae, or spores from one watershed to another [@nysdec2024].

The ecological damage caused by established invasive species takes many forms [@gentili2021]. Some are aggressive predators that decimate native prey populations before those populations have any chance to adapt [@mack2000]. Others are competitors that crowd out native plants or animals from food, space, and other resources they need to survive [@usda2024]. Over time, a successful invasion can fundamentally restructure an ecosystem — replacing native plant communities with monocultures, disrupting food webs, altering water cycles, and in the most severe cases, driving native species to local or even global extinction [@gentili2021; @mack2000]. Invasive species can also carry parasites and pathogens that spill over into native wildlife, compounding the damage furtherndefined \[@us

Beyond the ecological toll, invasive species impose enormous economic costs [@diagne2021]. Agricultural industries lose billions of dollars annually to crop damage from invasive pests and weeds [@diagne2021]. Forestry operations contend with invasive insects and diseases that kill commercially valuable trees [@mack2000]. Water infrastructure is disrupted by organisms like zebra mussels that clog intake pipes and alter aquatic chemistry [@usda2024]. And the cost of active management — detection, treatment, removal, and restoration — adds yet another layer to the financial burden [@diagne2021; @leung2002]. In densely populated and ecologically diverse states like New York, these costs are felt across sectors and communities [@nysdec2024].

New York is particularly exposed to invasive species pressure for several interconnected reasons [@nysdec2024]. It sits at the hub of one of the busiest transportation networks on the eastern seaboard, with major ports, airports, highways, and rail lines that move enormous volumes of goods and people [@nysdec2024]. Its landscape spans a remarkable range of ecosystems — coastal wetlands, deciduous forests, alpine zones, freshwater river systems — each of which represents a potential foothold for a different suite of invasive organisms [@nysdec2024]. And its long history of settlement, agriculture, and land use transformation has already altered many native habitats in ways that make them more vulnerable to invasion [@gentili2021]. Put all of that together, and it becomes clear why New York has invested significantly in building a statewide infrastructure for invasive species management [@nysdec2024].

### Background on NYS DEC Invasive Species Programs and the PRISM System

The New York State Department of Environmental Conservation (DEC) works with many groups — such as resource managers, non-profit organizations, industry workers, recreation groups, and other state agencies — to deal with invasive species [@nysdec2024]. To help with this work, New York created eight Partnerships for Regional Invasive Species Management (PRISMs) [@nysdec2024]. These PRISMs cover the entire state and help reduce the damage invasive species cause to New York's environment [@nysdec2016]. PRISMs were formed and funded under New York State law (Environmental Conservation Law 9-1705) [@nysdec2024].

PRISMs are groups of partners who work together to manage invasive species in different regions of New York [@nysdec2016]. They include people from federal and state agencies, environmental groups, scientists, industry, outdoor recreation, and local community members [@nysdec2024]. The DEC supports each PRISM through the Environmental Protection Fund, giving money to the organizations that host and coordinate them [@nysdec2024; @nysdec2016].

## PRISMs carry out many important tasks, including:

Since 2013, the DEC has contracted organizations to run all eight PRISMs across the state, which plan and coordinate regional invasive species management while bringing partners together statewide [@NYSDEC2016]. They survey and map invasions, detecting new threats early and responding quickly before species spread [@NYSDEC2016]. Control and removal projects are carried out alongside habitat restoration and monitoring after treatment [@NYSDEC2024]. Communities are engaged through education, outreach, volunteer training, and citizen science support through the Invasive Species Research Institute, with sightings logged to iMapInvasives [@NYSDEC2024].

## PRISMs deliver:

PRISMs provide comprehensive regional planning for invasive species management, ensuring coordinated strategies are in place across all eight regions of New York State [@NYSDEC2016]. They lead early detection and rapid response efforts, minimizing the establishment and spread of new invasions [@NYSDEC2016]. Eradication and control projects are executed on the ground, followed by long-term habitat restoration and monitoring to track recovery [@NYSDEC2024]. With DEC-supported educators, PRISMs deliver targeted education and outreach to communities, raising awareness and building local capacity [@NYSDEC2024]

The DEC is the main state agency responsible for environmental protection in New York, and invasive species management is part of its work [@nysdec2024]. Because invasive species spread across boundaries, New York created PRISMs so each region can coordinate its efforts [@nysdec2024].

The Lower Hudson PRISM (LHPRISM) covers the Lower Hudson Valley [@nysdec2024]. It uses trained staff and community volunteers who go into the field, record invasive species, help with treatment projects, and upload data to a shared statewide database [@nysdec2024]. This data is used for many environmental projects in the region [@finley2023].


### Information and Background on the NYS Invasive Species Lists

Not every non-native species that shows up in New York needs immediate action [@leung2002]. Deciding which species deserve attention — and how urgent that attention should be — requires a clear system for evaluating risk [@leung2002]. New York State uses two main systems to do this: a regulatory list managed by the Department of Environmental Conservation (DEC) and a tiered priority system created through collaboration between the PRISM network, state agencies, and the New York Natural Heritage Program (NYNHP) [@nysdec2024; @nynhp2024].

The regulatory list, created under the Environmental Conservation Law, sorts species into two groups based on how harmful they are and what restrictions are needed [@nysdec2024].

Prohibited species are considered so damaging, and so unnecessary for any legal use, that they cannot be possessed, sold, bought, transported, or introduced in New York at all [@nysdec2024]. There are no permits for these species — they are simply banned [@nysdec2024].

Regulated species are more complicated [@nysdec2024]. They may have some allowed uses, but there are still rules in place to prevent them from spreading [@nysdec2024].

Before a species is added to either category, experts review scientific information about its ecological impacts, how quickly it spreads, how feasible it is to control, and any economic or social considerations [@leung2002; @nysdec2024].

Alongside this legal framework, NYNHP created a tiering system to help PRISMs and other managers decide where to focus their limited time and resources [@nynhp2024]. Because New York has hundreds of invasive species, the tiering system uses a structured method that combines iMapInvasives data, invasiveness assessments, and expert input [@finley2023]. A species only receives a tier if it is considered to have high negative impacts [@nynhp2024]. Species that are present but not very harmful do not get tier assignments [@nynhp2024]. The tier definitions were finalized in 2017 through a statewide collaboration, giving all PRISMs a shared set of priorities [@nynhp2024; @finley2023].

These tiers can change over time [@nynhp2024]. They depend on accurate data, so new reports or successful management efforts can shift a species' tier [@finley2023]. This is why reporting new sightings to iMapInvasives is strongly encouraged — better data leads to better decisions [@nynhp2024]. For the species included in this report, their presence in the LHPRISM dataset means they have already been evaluated, found to be high-impact, and identified as priorities for management in the Lower Hudson region [@finley2023; @nysdec2024].

Understanding this background — the regulatory status, the tier level, and the management history of each species — is important for interpreting the data and understanding what it means for ongoing conservation work in this part of New York State [@nysdec2024; @nynhp2024].

## Region of study - Tanaka

The study area for this project encompasses the Lower Hudson region, as delineated by the Lower Hudson Partnership for Regional Invasive Species Management (LH PRISM). Established as one of eight regional invasive species management groups across New York State, LH PRISM coordinates collaborative efforts among government agencies, conservation organizations, and landowners to monitor and manage invasive species at a regional scale [@lhprism2024]. The region includes Westchester, Dutchess, Putnam, Rockland, Orange, and Ulster counties, extending southward into the five boroughs of New York City [@lhprism2024; @nysdec2024].

Geographically, the region is anchored by the Hudson River estuary, a highly dynamic tidal system in which freshwater inputs from inland watersheds interact with saltwater intrusion from the Atlantic Ocean. This estuarine gradient sustains a broad array of habitat types, including tidal wetlands, riparian corridors, freshwater marshes, and open aquatic environments, collectively rendering the Lower Hudson one of the most ecologically productive systems in the northeastern United States [@dec_aquatic2024]. The region also functions as a critical landscape transition zone, bridging the heavily urbanized New York City metropolitan area with the more rural and forested landscapes of the mid-Hudson Valley. This gradient of land use intensities is particularly significant in the context of biological invasions, as anthropogenic disturbance is well-documented as a primary driver of invasive species establishment and spread [@cornell2024; @nysdec2024].

### Climate

The Lower Hudson region is characterized by a humid temperate climate with four well-defined seasons. Summers are warm and humid, winters are cold with periodic snowfall, and precipitation is distributed relatively uniformly throughout the year, averaging approximately 40 to 50 inches annually [@noaa2024]. These climatic conditions support a wide diversity of native vegetation communities while simultaneously creating favorable thermal and moisture regimes for the establishment of non-native invasive species [@kunkel2013; @noaa2024]. Notably, climate projections for the northeastern United States indicate continued warming trends and increasing precipitation variability, conditions that are expected to further facilitate the range expansion of invasive species into previously unsuitable habitats [@kunkel2013].

### Ecology

The Lower Hudson region falls within the Eastern Temperate Forest biome and supports a mosaic of ecological communities, including mixed deciduous forests, freshwater wetlands, floodplain systems, and estuarine habitats [@nyheritage2021]. Dominant forest canopy species include oak (*Quercus* spp.), maple (*Acer* spp.), and hickory (*Carya* spp.), which provide structural complexity and support high levels of faunal biodiversity across the landscape [@nyheritage2021; @cornell2024]. The region's network of river systems and wetlands further contributes to its ecological significance, serving as corridors for both native wildlife and, unfortunately, the dispersal of invasive plant species [@dec_aquatic2024].

The Lower Hudson region presents an ideal study area for analyzing invasive species distribution patterns given its unique combination of ecological diversity, landscape heterogeneity, and anthropogenic pressure. The juxtaposition of urban and natural land cover types creates conditions highly conducive to invasive species introduction, as disturbance-adapted non-native plants readily exploit habitat edges, degraded areas, and human-modified landscapes [@lhprism2024; @nysdec2024]. The dataset employed in this study, compiled through field surveys conducted by the Invasives Strike Force under the coordination of LH PRISM, provides spatially explicit records of invasive plant populations across a broad range of property types within this region. Species identifications in the dataset are standardized using codes from the USDA PLANTS Database, ensuring taxonomic consistency across observation records [@usda2024]. Together, the ecological characteristics of the Lower Hudson region and the availability of a robust regional dataset make this an appropriate and well-suited geographic context for the analyses undertaken in this project.

## A map showing the representation of the Region of the Study is shown in the Spatial Analysis Section

**References**

Cornell University Department of Natural Resources and the Environment. (2024). *Conservation priorities in the watershed*. <https://hudson.dnr.cals.cornell.edu/natural-areas-biodiversity/conservation-priorities-watershed>

Kunkel, K. E., & others. (2013). *Regional climate trends and scenarios for the U.S. National Climate Assessment: Part 1. Climate of the Northeast U.S.* NOAA Technical Report NESDIS. <https://scenarios.globalchange.gov/sites/default/files/NOAA_NESDIS_Tech_Report_142-1-Climate_of_the_Northeast_US.pdf>

Lower Hudson Partnership for Regional Invasive Species Management. (2024). *About NYS PRISM network and the Lower Hudson*. <https://www.lhprism.org/about-us/>

National Centers for Environmental Information. (2024). *U.S. climate normals*. <https://www.ncei.noaa.gov/products/land-based-station/us-climate-normals>

New York Natural Heritage Program. (2021). *Ecological communities of New York State*. <https://guides.nynhp.org/ecological-communities/>

New York State Department of Environmental Conservation. (2024). *Aquatic habitats of the Hudson River Estuary*. <https://dec.ny.gov/nature/waterbodies/oceans-estuaries/hudson-river-estuary-program/aquatic-habitats>

New York State Department of Environmental Conservation. (2024). *Invasive species*. <https://dec.ny.gov/nature/animals-fish-plants/invasive-species>

USDA Natural Resources Conservation Service. (2024). *PLANTS database*. <https://plants.usda.gov>

## Overview of the LHPRISM data set

## Dataset overview

The dataset is an environmental monitoring record compiled by the Lower Hudson Partnership for Regional Invasive Species Management (LHPRISM). This organization operates within the broader NY-NJ region. Records span from 1992 to 2022, meaning we do not have a static snapshot of the region, but instead an archive of ongoing invasive management. This dataset is concerned with invasive plant species and the coordinated effort to control and eliminate such species from targeted sites across the northeast. Both the collection period and geographical range are broad, with records drawn from over a dozen counties. Sites range from parks, wildlife sanctuaries, golf courses, and nature preserves. Such a swath reflects the extent of ecological damage caused by invasive species.

The dataset comprises just under 1,050 distinct sites that are each assigned a unique code and linked to contact information. The ecological content is split into populations and visits. The Populations tab documents 2,127 individual invasive species infestations that are assigned a specific property code and plant species code, georeferenced with latitude and longitude coordinates, and assigned a current treatment status. These statuses range from "Untreated" to "Eradicated", with intermediate stages including "Partially Treated," "Completely Treated," "No Plants 1 Year," and "No Plants 2 Years." This graduated classification system is central to the dataset's design, as it is precisely this progression, or lack of progression, that the dataset was built to track.

```{r}
#| label: tbl-metadata
#| tbl-cap: "LHPRISM Invasive Strikeforce Repeat Visits Metadata"

#----------------------------------------------------------
# METADATA TABLE
#----------------------------------------------------------
metadata <- read_csv(
  paste0(proj_path, "/data/metadata.csv"),
  show_col_types = FALSE
) %>% clean_names()

metadata$variable_name_clean <- make_clean_names(metadata$variable_name)

missing_vec <- colMeans(is.na(lhprism_data))

metadata$percent_missing <- round(
  missing_vec[metadata$variable_name_clean] * 100,
  2
)

metadata_display <- metadata %>%
  select(
    variable_name_clean,
    description,
    units,
    percent_missing
  ) %>%
  rename(
    "Variable Name" = variable_name_clean,
    "Description" = description,
    "Units" = units,
    "Percent Missing Data" = percent_missing
  )

kable(metadata_display, escape = TRUE, format = "simple")
```

The species catalog is diverse, containing `{r} length(unique(lhprism_data$species_code))` unique species across the region. 
Some of the more predominant invasives include multiflora rose, Japanese knotweed, and mile-a-minute vine. 
The Visit sheets contain nearly 3,830 records representing the field backbone of the dataset. 
Each visit log captures the site surveyed, the date of the survey, the person performing the survey, and the conditions of the site. 
Variables record plant count, percent cover, phenological stage, the size of the area treated, and herbicide application specifics such as product name, concentration, and method used. 
This operational data is where the utility for evaluation arises, since it also tracks each site over time. 
This dataset is a collaboration between at least 18 distinct organizations. 
That collaboration is a key feature of the dataset and introduces meaningful variation in the data, as it may reflect what fields were prioritized and the consistency of terminology used. 
For example, the appearance of both “2-Completely Treated” and “Completely Treated 2” as distinct entries shows that standardization was not always consistent.

Prior to data analysis, our team completed a data cleaning procedure on OpenRefine to ensure consistency across entry variations. Missingness in variables pertaining to plant count and cover may carry more weight than treatment method variables, since every record likely did not undergo every possible treatment, but every record should have had a baseline count value to assess treatment efficiency. Variables with higher percentages of missing data may still be useful, but may also lead to gaps in final analysis.

Uncertainty caused by gaps in this dataset primarily originates from a lack of direct communication with LHPRISM staff. Our ability to confirm procedural methods has been limited, including procedure design, implementation, and standardization across individuals visiting these sites. Though not the primary focus of our analysis, several assumptions were made for categorical variables:

- Variables addressing percent buckets likely follow traditional ecological methods in which observers categorize visual estimates into 0–5, 6–25, 26–50, 51–75, and 76–100 percent classes.
- Distribution, native, and non_nat variables likely followed the same concept; observers probably did rapid visual assessments for category classification rather than taking numerical measures.
- Phenology and age_size were likely classified by assessing visible physical characteristics and size of vegetation.

It is unknown how sites were selected and defined. The data does not show whether sites were chosen randomly or selected because of high invasive presence. It is also possible that sites were chosen opportunistically for accessibility or to support quick evaluation of treatment efficiency. Not knowing how locations were chosen introduces probable bias. For example, a dataset selected because of denser invasive populations could skew results toward harder-to-treat sites, which would make treatment efficacy appear lower than it would across a random selection of sites. Multiple observers visited unique sites over time, which may affect consistency in visual assessments of percent_cover or similar non-standardized variables. Additionally, there is no information on whether observers were responding to the same sub-area within one site, as search_sqft and gross_sqft differ across recordings under one site code. We are unsure whether these inconsistencies reflect opportunity to visit or intentional collaboration. Variation in data entry across individual observers has the potential to affect analysis, since gaps may come from differences in technique rather than true ecological response.

```{r}
#| label: tbl-site-history
#| tbl-cap: Visit history and treatment status for site ATNY_SAGL12_001

# Filter for the site shown in the screenshot
site_history <- lhprism_data %>%
  filter(site_code == "ATNY_SAGL12_001") %>%
  select(site_code, visit_date, visit_status_char) %>%
  mutate(
    visit_date = as.Date(visit_date),
    date_display = format(visit_date, "%m/%d/%Y")) %>%
  arrange(visit_date) %>%
  select(date_display, visit_status_char)

# Display as a labeled figure/table
kable(site_history)
```

Categories in visit_status introduce a heavier uncertainty because we are unsure if observers were prompted to record the site’s current treatment condition or to only document any work they completed at the time of visit. The sequence of entries for a site on the Appalachian Trail regarding sticky sage (Fig. #) leads with a “completely treated” entry but when visited later by a different observer, is listed as “untreated”.  This pattern suggests the latter, in which observers recorded their own actions rather than combining a site’s treatment history. If true, the reversion as listed could affect analyses about the time it takes to get an invasive species to eradication: a site listed as “untreated” because no new work was performed could lengthen the timeline between treatment stages and subsequently final success. A lack of clarity from LHPRISM staff on how observers were trained to apply categories, especially in a variable that is critical in evaluating treatment outcomes, leaves us unsure whether the data represents ecological changes or inconsistencies in procedure across observers.


## Follow-up visits per site

```{r visits-site}
plot_visits <- lhprism_data %>%
  group_by(site) %>%
  summarise(visit_count = n())

ggplot(plot_visits, aes(x = site, y = visit_count)) +
  geom_bar(stat = "identity") +
  labs(title = "Follow-up Visits per Plot", x = "Site", y = "Number of Visits") +
  theme(axis.text.x = element_blank())

```

From this figure, it is clear that the frequency of visits across sites is highly uneven, with a small number of sites accounting for a disproportionate share of the total visits. This skew is important to note as we consider treatment outcomes, because heavily visited sites may show stronger results simply due to the volume of intervention.

## Treatment counts

```{r treatment-counts}
treatment_counts <- data.frame(
  Treatment = c("PU", "FO", "CO", "CS", "BB", "HS"),
  Count = c(
    sum(!is.na(lhprism_data$pu_count)),
    sum(!is.na(lhprism_data$fo_count)),
    sum(!is.na(lhprism_data$co_count)),
    sum(!is.na(lhprism_data$cs_count)),
    sum(!is.na(lhprism_data$bb_count)),
    sum(!is.na(lhprism_data$hs_count))
  )
)

ggplot(treatment_counts, aes(x = Treatment, y = Count)) +
  geom_col(fill = "#5AAE61") +
  labs(
    title = "Count of Treatment Types",
    x = "Treatment",
    y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold")
  )
```

Pull by hand (PU) is by far the most common treatment method across the dataset. Cut stomp (CS) and basal bark (BB) applications were also frequent, while hack and squirt was rarely recorded. This distribution likely reflects both species composition and practical constraints in the field; hand pulling is easier and cheaper than chemical application. The limited use of herbicide may also reflect site restrictions, since chemical application at some preserves may be more tightly regulated.

## Visits by species

```{r visits-species}
species_visits <- lhprism_data %>%
  group_by(species_code) %>%
  summarise(visit_count = n(), .groups = "drop") %>%
  arrange(desc(visit_count))

ggplot(species_visits, aes(x = reorder(species_code, -visit_count), y = visit_count)) +
  geom_col(fill = "#F4A582") +
  labs(
    title = "Visits by Species",
    x = "Species",
    y = "Number of Visits"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_blank()
  )
```

The distribution of visits by species is strongly right-skewed, with a few key species accounting for the majority of recorded visits. This pattern is consistent with what we know about the region: a small number of aggressive invasive species dominate the landscape and demand repeated intervention. The long tail may reflect either limited spread or successful management, since some species were clearly far more common than others in total visit numbers.

## Plot visit frequency

```{r visit-buckets}
plot_visit_counts <- lhprism_data %>%
  group_by(site_code) %>%
  summarise(visit_count = n_distinct(visit_date), .groups = "drop")

plot_visit_bucketed <- plot_visit_counts %>%
  mutate(
    visit_bucket = case_when(
      visit_count == 1 ~ "1 Visit\n(No Follow-up)",
      visit_count >= 2 & visit_count <= 3 ~ "2–3 Visits",
      visit_count >= 4 & visit_count <= 5 ~ "4–5 Visits",
      visit_count >= 6 & visit_count <= 10 ~ "6–10 Visits",
      visit_count > 10 ~ "10+ Visits"
    ),
    visit_bucket = factor(
      visit_bucket,
      levels = c(
        "1 Visit\n(No Follow-up)", "2–3 Visits", "4–5 Visits",
        "6–10 Visits", "10+ Visits"
      )
    )
  ) %>%
  count(visit_bucket, name = "n_plots")

no_followup <- plot_visit_bucketed$n_plots[plot_visit_bucketed$visit_bucket == "1 Visit\n(No Follow-up)"]
total_plots <- sum(plot_visit_bucketed$n_plots)

ggplot(plot_visit_bucketed, aes(x = visit_bucket, y = n_plots)) +
  geom_col(fill = "#2C7BB6", width = 0.65) +
  geom_text(aes(label = n_plots), vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Distribution of Follow-up Visits Across Plots",
    subtitle = paste0(
      no_followup, " of ", total_plots, " plots (",
      round(100 * no_followup / total_plots, 1), "%) received no follow-up"
    ),
    x = "Visit Frequency",
    y = "Number of Plots"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    panel.grid.major.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

This figure shows that follow-up effort was concentrated in a smaller subset of plots. Many plots were visited only once, meaning they received no follow-up, while a smaller number of plots received repeated attention. That pattern suggests that the monitoring program focused intensively on certain locations while leaving many others with limited repeat observation.

Categories in visit_status introduce a heavier uncertainty because we are unsure whether observers were prompted to record the site’s current treatment condition or only document work completed during that specific visit. The sequence of entries for a site on the Appalachian Trail regarding sticky sage suggests this ambiguity: one record lists the plot as completely treated, but a later visit by a different observer lists it as untreated. This pattern suggests that observers may have recorded their own actions rather than a full treatment history. If so, reversions in status could distort analyses about the time required to move a species toward eradication. A site listed as untreated because no new work was performed could appear to regress in the treatment timeline, even if the actual ecological condition had not changed.

The overall purpose of the dataset is to evaluate the effectiveness of treatment interventions over time and to estimate how long it takes to move a species toward eradication. By tracking multiple sites over multiple years, the dataset highlights important questions: Which species respond best to intervention? Which species persist despite repeated treatment? Under what conditions are populations most effectively reduced or eradicated? Beyond individual population outcomes, the dataset is intended to surface broader patterns, trends in species spread, shifts in treatment methods, and regional variation in management success that can inform future resource allocation and strategic planning for invasive species control across the lower Hudson Valley.

-   **———————————————————————————————————————**

## Spatial distribution of the data

The Leaflet JavaScript library was used to create two maps for this dataset: one depicting the exact locations of each population that was treated and another displaying their treatment statuses. Map one displays the properties at which the invasive species in this project were observed.Users can interact with it by clicking on each data point to see the name of its corresponding property. Map two describes the extent to which the populations at each location were treated using a color coordination scheme. These maps provide a tidy visual aid for users by offering a straightforward presentation with a useful interface that allows users to learn more about any data point they choose.

Leaflet relies heavily on the pipe function to allow users to add layers to maps. To create the first map, users will load four packages: leaflet, readxl, tidyverse, and here. After loading these packages and reading in the data, the pipe operator is used with the filter function to eliminate NA values from the data. Then, the "\$" extraction operator is used to comb through the data, isolating points by latitude, longitude, and property name as separated by columns in Excel.

In the LHPRISM data set, there is a point whose longitude included an errant zero, putting it well out of range of the other data points. This error was corrected, incorporating the point back into the range of the others. Once the data is formatted appropriately, it can be displayed by creating a data frame with the isolated information related to location. Once this step is complete, the leaflet function itself is used to create the map object. The final step is to one again use the pipe operator, addTiles, and addMarkers functions to place the points on the map.

The second map uses a color gradient to classify points based on how far along the invasive species at these locations are in their treatment. The labels used are "Untreated," "Partially Treated," "Completely Treated," "No plants for 1 year," "No plants for 2 years," and "Eradicated." As with map one, the pipe operation is used with the addTiles and addCircleMarkers functions. These functions are used to assign different colors, along a spectrum, to pointes as determined by the status of their treatment. Untreated sites are grey, partially treated sites are orange-red, completely treated sites are yellow, sites with no plants for a year are yellow-green, sites with no plants for two years are green, and sites that have been eradicated of invasive plants are represented with lime green. This gradient provides a display that is easy for readers to understand and follow along with.

- Leaflet maps of populations:

```{r map_1}
# Patch to the Excel sheet
excel_file <- paste0(proj_path, "/data/Pace Data Project Invasives Strike Force Records.xlsx")

# Read in sheet 2 of the Excel spreadsheet (populations)
lhprism_pops <- read_excel(path = excel_file, sheet = "Populations", 
                           col_types = c("text", "text", "text", "text", 
                                         "text", "text", "text", "numeric", 
                                         "text", "text", "text", "text", "text", 
                                         "text", "text", "text", "text", "text", 
                                         "text", "text", "text", "text"))

lhprism_pops$LAT <- as.numeric(trimws(lhprism_pops$LAT))
lhprism_pops$LONG <- as.numeric(trimws(lhprism_pops$LONG))

# Remove NA values
lhprism_pops <- lhprism_pops %>%
  filter(!is.na(LONG))

# Isolate Latitude, Longitude, Population status and Property name
location_name <- lhprism_pops$PROPERTY_NAME
location_lat <- lhprism_pops$LAT
location_lng <- lhprism_pops$LONG

# Fixed data point that in the ocean due to incorrect data entry.
# Latitude of 40.0975937 changed to 40.975937 (Rye Nature Center)
location_lat[234] <- 40.975937

# Build data frame with latitude and longitude values
df1 <- data.frame(name = location_name,
                 lat = location_lat,
                 lng = location_lng)
# Create leaflet map object
m1 <- leaflet()

# Generate points for each population's coordinates
# Click on each point to see the Property Name
m1 %>%
  addTiles() %>%
  addMarkers(data = df1, popup = location_name)


```

Map 1: Locations of each treatment population. If viewing the interactive map, clicking on each point will bring up its associated property name (e.g. Rye Nature Center).

```{r map_2}
## To-Do - need to think about how to display this through time.

# Fix status text: ""1-Partially Treated" instead of "1- Partially Treated""
lhprism_pops$POP_STATUS[3] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[7] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[10] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[15] <- "1- Partially Treated"

# Fix status text: ""2-Completely Treated" instead of "2- Completely Treated""
lhprism_pops$POP_STATUS[1] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[4] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[5] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[6] <- "2- Completely Treated"

# Build new data frames for each status category
stat_0 <- filter(lhprism_pops, POP_STATUS == "0- Untreated")
stat_1 <- filter(lhprism_pops, POP_STATUS == "1- Partially Treated")
stat_2 <- filter(lhprism_pops, POP_STATUS == "2- Completely Treated")
stat_3 <- filter(lhprism_pops, POP_STATUS == "3- No Plants 1 Yr")
stat_4 <- filter(lhprism_pops, POP_STATUS == "4- No Plants 2 Yrs")
stat_5 <- filter(lhprism_pops, POP_STATUS == "5- Eradicated")

# Fixing the same incorrectly entered latitude value (Rye Nature Center)
stat_1$LAT[37] <- 40.975937

# Generate points colored by visit status
m2 <- leaflet()

m2 %>%
  addTiles() %>%
  addCircleMarkers(data = stat_0,
                   lng=~stat_0$LONG,
                   lat=~stat_0$LAT,
                   popup=~paste0("Site Name: ",stat_0$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_0$POP_STATUS),
                   color="grey",
                   radius = 4,
                   opacity = 0.7,
                   fillColor = "white",
                   group= "0- Untreated"
  ) %>% 
  addCircleMarkers(data = stat_1,
                   lng=~stat_1$LONG,
                   lat=~stat_1$LAT,
                   popup=~paste0("Site Name: ",stat_1$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_1$POP_STATUS),
                   color="orangered",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "1- Partially Treated"
  ) %>% 
  addCircleMarkers(data = stat_2,
                   lng=~stat_2$LONG,
                   lat=~stat_2$LAT,
                   popup=~paste0("Site Name: ",stat_2$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_2$POP_STATUS),
                   color="yellow",
                   radius = 4,
                   opacity = 0.6,
                   fillColor = "white",
                   group= "2- Completely Treated"
  ) %>% 
  addCircleMarkers(data = stat_3,
                   lng=~stat_3$LONG,
                   lat=~stat_3$LAT,
                   popup=~paste0("Site Name: ",stat_3$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_3$POP_STATUS),
                   color="yellowgreen",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "3- No Plants 1 Yr"
  ) %>% 
  addCircleMarkers(data = stat_4,
                   lng=~stat_4$LONG,
                   lat=~stat_4$LAT,
                   popup=~paste0("Site Name: ",stat_4$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_4$POP_STATUS),
                   color="green",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "4- No Plants 2 Yr"
  ) %>% 
  addCircleMarkers(data = stat_5,
                   lng=~stat_5$LONG,
                   lat=~stat_5$LAT,
                   popup=~paste0("Site Name: ",stat_5$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_5$POP_STATUS),
                   color="limegreen",
                   radius = 4,
                   opacity = 0.9,
                   fillColor = "white",
                   group= "5- Eradicated"
  ) %>%
  addLegend(
    position = "topleft",
    color=c("grey", "orangered", "yellow", "yellowgreen", "green", "limegreen"),
    labels=c("Untreated","Partially Treated","Completely Treated","No plants for 1 year", "No plants for 2 years", "Eradicated"),
    opacity = 0.9
  )
```

Map 2: Locations of each treatment population, colored by status of treatment (e.g. "Untreated", "Partially Treated"). If viewing the interactive map, clicking on each point will bring up its associated property name and visit status.

## Analysis of time / treatment effort required to eradication

- how many times did they need to visit a plot before eradication: look at total number of visits, and population numbers

- Description of the results of this analysis

- How Many site codes have been Eradicated (5)

- This will be done on a subset of species

- How did we choose these species?

- these species were chosen in part because LHPRISM believes eradication is possible and for some of these species like Giant Hogweed, specific traits that are hazardous to humans were also taken into account.

- Brief overview of ecology of these particular species

- Description of the methods used in this analysis

- Description of the results of this analysis

- accelerated growth, previous exploratory analysis file - HEMA17 - Giant Hogweed - PUMO - Kudzu - CYSC4 - Scotch Broom - SAGL12 - Sticky Sage - CYSC4, SAGL12, LECU, PLEIO, POPE10, RULA, AREL8, HEMA17, ACAR10, VISI, COIN21, PUMO, ARIT, ILCR2, PHVI81, SYPA12, VIDI80, COKO2, PHVI81, SYPA12, AKQU, AREL8, SPJA, WISI, PHVI81, BRPA4, MISI, RAFIB, BRSY, HUJA, HEHE, FRAL4, WISTE - ALL EXTERMINATED

- Brief overview of ecology of these particular species (add known methods of treatment, previous recommendation)

- HEMA17 - Giant Hogweed (Heracleum mantegazzianum)

- native to Eastern Europe/Western Asia, highly invasive in NE USA

- thrives in rich, moist soil, prefers sunlight and tolerant of partial shade

- found in river banks, streams, floodplains, roadsides + trafficked areas

- biennial lifecycle, plants 2+ yrs old 10-15ft tall

- 1 plant can produce 20,000-50,000 seeds, spreading by water/wind and human activity

- forms dense monocultures, increases erosion risk due to dying back in winter, leaves soil exposed along riverbanks

- phototoxic sap - sever skin burns with sun exposure

- treatment recommendation: chemical treatment in early spring, mechanical removal for small infestations

- PUMO - Kudzu

- native to East Asia, widespread in SE USA and migrating N

- thrives on roadside, forest edges, warm and humid climates, prefers full sun, climbs to canopy

- perennial vine, grows up to 1ft/day in optimal conditions

- spreads vegetatively, rooting at nodes

- forms dense monocultures, causes tree death by shading and structural stress

- alters nutrient cycle by nitrogen fixation

- treatment complex, requires multi-modal approach over multiple years

- CYSC4 - Scotch Broom

- Native to Europe, widespread invasiveness in PNW, California, and East Coast USA

- prefers sandy soil and full sun, found on roadsides, logged forests, grassland

- perennial shrub, lives up to 10-15 yrs

- 1 plant produces 1,000s of seeds

- dense thickets, reduces habitat quality for native species, alters nutrient cycle by nitrogen fixation

- seeds viable for 30+ yrs in soil makes eradication difficult

- hand pulling for small infestations, chemical spray (glyphosate, triclopyr, basal bark treatment, stem cutting, foliar spray

- SAGL12 - Sticky Sage

- native to Europe and West Asia

- prefers moist, well-drained soil, found in woodlands, shaded slopes, forest edges

- perennial herb, blooms late summer

- low ecological impact, coexists with native species

- known treatment: handpulling for young pop

- Description of the methods used in this analysis

- change in plant count over consecutive visits

- visit_date \* visit_status_num

```{r status_time}
##created a table to see how long does it take each species to move from one status to the next, sorting by site
status_timing <- lhprism_data %>%
  group_by(species_code, site, visit_status_num) %>%
  summarise(first_date = min(visit_date), .groups = "drop") %>%
  arrange(species_code, site, visit_status_num)

##this does not account for status relapses (yet) as it only searches for the earliest date that each visit_status_num was recorded
```

- maybe see if there are any patterns when sorting by herbicide/treatment method?

- how many times did they need to visit a plot before eradication: look at total number of visits, and population numbers

```{r hema17_status_time}
##HEMA17 population by site and visit_status_num
hogweed_pop <- lhprism_data %>%
  filter(species_code == "HEMA17") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#plotting visit_status_num across time
ggplot(data = hogweed_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()
```

```{r pumo_status_time}
##PUMO population by site and visit_status_num
jknotweed_pop <- lhprism_data %>%
  filter(species_code == "PUMO") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#PUMO visit_status_num over time
ggplot(data = jknotweed_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()

##plant_count at each status level
##days/visits until new status, also days until relapse?

#select(site, visit_date, visit_status_num, plant_count)

```

```{r kudo_status_time}
##KUDO population by site and visit_status_num
jknotweed_pop <- lhprism_data %>%
  filter(species_code == "KUDO") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)
```

```{r cycsc4_status_time}
##CYSC4 population by site and visit_status_num
scotchbroom_pop <- lhprism_data %>%
  filter(species_code == "CYSC4") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)
```

```{r sagl12_status_time}
##SAGL12 population by site and visit_status_num
stickysage_pop <- lhprism_data %>%
  filter(species_code == "SAGL12") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)

#PUMO visit_status_num over time
ggplot(data = stickysage_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()
```

- Description of the results of this analysis \`\`\`

- How Many site codes have been Eradicated (5)

```{r}
lhprism_treated <-
  lhprism_data %>%
  filter(grepl("5", visit_status_num)) %>%
  mutate(Eradicated = ifelse(visit_status_num == "5", 1, 0))

```

- Count of visits 0-5 status

```{r}
Visit_num_vs_site_code <-
  lhprism_data %>%
  group_by(visit_status_num, site_code) %>%
  count() %>%
  filter(!is.na(site_code)) %>%
  pivot_wider(names_from = site_code, values_from = n)
lhprism_treated
```
-  The three species found to have been fully eradicated at certain sites are Scotch Broom (CYSC4), Giant Hogweed (HEMA17), and Kudzu (PUMO). Linear regression graphs were calculated for each.

-   visit_status_num \* visit_date, examining how long it takes to progress to the next status for 4 target species

- visit_status_num \* visit_date, examining how long it takes to progress to the next status for 4 target species

- HEMA17

```{r}
##HEMA17 population by site and visit_status_num
hogweed_status <- lhprism_data %>%
filter(species_code == "HEMA17") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#plotting visit_status_num across time
ggplot(data = hogweed_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()

##plant_count at each status level?
##days/visits until new status, also days until relapse?


hogweed_status_vs_visit <-
hogweed_treated %>%
group_by(`Site Code`) %>%
summarise(total_visits = n(), 
          max_invasion = max(GROSS_SQFT, na.rm = TRUE),
          final_status = max(VISIT_STATUS_NUM))

hogweed_status_vs_visit$max_invasion[which(hogweed_status_vs_visit$final_status == 0)] <- NA 

  ggplot(data = hogweed_status_vs_visit, aes(y=final_status, x = total_visits, color= `Site Code`)) + 
  geom_point() + 
  ylim(c(0,5)) +
  xlim(c(0,10))

  ggplot(data = hogweed_status_vs_visit, aes(y=final_status, x = max_invasion)) + 
  geom_point() +
  geom_smooth(method = "lm")
```
```

- PUMO

```{r}
##PUMO population by site and visit_status_num
jknotweed_status <- lhprism_data %>%
filter(species_code == "PUMO") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#PUMO visit_status_num over time
ggplot(data = jknotweed_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()

jknotweed_status_vs_visit <-
jknotweed_treated %>%
group_by(`Site Code`) %>%
summarise(total_visits = n(), 
          max_invasion = max(GROSS_SQFT, na.rm = TRUE),
          final_status = max(VISIT_STATUS_NUM)) 

jknotweed_status_vs_visit$max_invasion[which(jknotweed_status_vs_visit$final_status == 0)] <- NA 

  ggplot(data = jknotweed_status_vs_visit, aes(y=final_status, x = total_visits, color= `Site Code`)) + 
  geom_point() + 
  ylim(c(0,5)) +
  xlim(c(0,15))

ggplot(data = jknotweed_status_vs_visit, aes(y=final_status, x = max_invasion)) + 
  geom_point() +
  geom_smooth(method = "lm")

```

- CYSC4

```{r}
##CYSC4 population by site and visit_status_num
scotchbroom_status <- lhprism_data %>%
filter(species_code == "CYSC4") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count)

#CYSC4 visit_status_num over time
ggplot(data = scotchbroom_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()
```

The only scotch broom site that was fully eradicated was HARR_CYSC4_040 with 4 total visits
```{r}
scotchbroom_status_vs_visit <-
scotchbroom_treated %>%
group_by(`Site Code`) %>% 
          summarise(total_visits = n(), 
          max_invasion = max(GROSS_SQFT, na.rm = TRUE),
          final_status = max(VISIT_STATUS_NUM))

scotchbroom_status_vs_visit$max_invasion[which(scotchbroom_status_vs_visit$final_status == 0)] <- NA 


ggplot(data = scotchbroom_status_vs_visit, aes(y = final_status, x = total_visits, color= `Site Code`)) + 
  geom_point() + 
  xlim(c(0,25))

ggplot(data = scotchbroom_status_vs_visit, aes(y=final_status, x = max_invasion)) + 
  geom_point() +
  geom_smooth(method = "lm")


```
-   SAGL12

```{r}
##SAGL12 population by site and visit_status_num
stickysage_status <- lhprism_data %>%
filter(species_code == "SAGL12") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count)

#PUMO visit_status_num over time
ggplot(data = stickysage_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()
```

- Description of the results of this analysis

- HEMA17

```{r}
```

- PUMO

```{r}
```

- CYSC4

```{r}
```

- SAGL12

```{r}
```

## Conclusions and recommendations
