---
title: "LHPRISM Exploratory Analysis"
format: docx
editor: visual
---

## Overview

This document goes through various exploratory analyses of the LHPRISM post-treatment invasive species survey data set.

## Data cleaning

This step was completed using OpenRefine. The JSON code for all steps can be found in the file `data/lhprism_openrefine_cleaning.json`. The key data cleaning steps and assumptions taken are listed below.

-   LIST KEY DATA CLEANING STEPS AND ASSUMPTIONS
-   MAKE SURE TO POINT OUT ASSUMPTIONS ASSOCIATED WITH PERCENTAGE CATEGORIES

## Reading in the data

The data are provide to us as an Excel file with multiple sheets. We are most interested in the sheet named "Visits". We will use the `readxl` package to read in the data into our work environment. But first, we will use the `pacman` package to load, and install if needed, any other packages. If this is the first time you are using `pacman`, remember that you first need to install this package.

```{r}
## Get packages
library(pacman)
pacman::p_load(here, tidyverse, janitor, lubridate)

## Get the project level directory path
proj_path <- here()

## Define a color blind friendly palette
## These are from - http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

In this code chunk, we will read in the data file. You may have to edit this call, depending on where you have saved your version of the Excel file.

```{r}
## Read in the data file
lhprism_data <- read_csv(paste0(proj_path, "/data/LHPRISM-Data.csv"), 
    col_types = cols(`iMap Treat` = col_number(), 
        GROSS_TREAT_SQFT = col_number())) %>%
  clean_names()

## Edit the order of levels for some of the variables
lhprism_data <- lhprism_data %>%
  mutate(native = factor(native, levels = c("low", "medium", "high")),
         non_nat = factor(non_nat, levels = c("low", "medium", "high")))

## Create a new species code column
lhprism_data$species_code <- str_extract(lhprism_data$site_code, "(?<=_)[^_]+(?=_)") 
## Remove plant column
lhprism_data$plant <- NULL

## Create a new site column using the species_code column
lhprism_data$site <- str_extract(lhprism_data$site_code, "^[^_]+")

## Remove all rows that have NA for site
lhprism_data <- lhprism_data %>%
  filter(!is.na(site))

```

## Figures

```{r}
#| label: fig-native-vs-nonnative-coverage
#| fig-cap: "Relationship between native vs non-native coverage categories"
#| fig-width: 6


lh_sub <- lhprism_data %>%
  select(native, non_nat)

lh_sub_long <- 
  data.frame(native_vs_nonnative = rep(c("native", "non_nat"), each = nrow(lh_sub)),
             coverage = c(lh_sub$native, lh_sub$non_nat))

lh_sub_long %>%
  filter(!is.na(coverage)) %>%
  ggplot(., aes(x = coverage, fill = native_vs_nonnative)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = cbPalette) +
  xlab("Coverage Category") +
  ylab("Visit Count") +
  labs(fill = "Coverage Type") +
  theme_minimal()
```

Let's make a table that goes along with this.

```{r}
#| label: tbl-native-vs-nonnative-coverage-table

native_vs_nonnative_cover <-
  lhprism_data %>% 
  group_by(native, non_nat) %>%
  count() %>%
  filter(!is.na(native)) %>%
  pivot_wider(names_from = non_nat, values_from = n, values_fill = 0)

knitr::kable(native_vs_nonnative_cover,
             caption = "Counts of native vs non-native coverage categories. Note that rows are native classificaitons and columns are non-native classifications.")

## Perform a chi-squared analysis
native_vs_nonnative_cover_chisq <-
  chisq.test(native_vs_nonnative_cover[2:4])

native_vs_nonnative_cover_chisq
native_vs_nonnative_cover_chisq$residuals
```

### Histogram of visit date

```{r}
ggplot(data = lhprism_data, aes(x = visit_date)) +
  geom_histogram()
```

### Histogram of pu_count

```{r}
ggplot(data = lhprism_data, aes(x = pu_count)) +
  geom_histogram() +
  scale_x_log10()
```

### Relationship between gross square feet searched vs treated

```{r}
ggplot(data = lhprism_data, aes(x = gross_sqft, y = gross_treat_sqft)) +
  geom_point()
```

## Tables

Let's make some tables.

```{r}
#| label: tbl-visit-status-table
#| fig-cap: "Counts of visit status categories. Note that the 'NA' category is for visits where the visit status was not recorded."

lhprism_data %>%
  group_by(visit_status_char) %>%
  count() %>%
  filter(!is.na(visit_status_char)) %>%
  arrange(desc(n)) %>%
  rename(tot_count = n)

lhprism_data %>%
  group_by(visit_status_num) %>%
  count() 

lhprism_data %>%
  group_by(percent_treated) %>%
  count()

lhprism_data %>%
  group_by(distribution) %>%
  count()

lhprism_data %>%
  count(month(visit_date, label = TRUE)) %>%
  filter(!is.na(`month(visit_date, label = TRUE)`)) %>%
  rename(Month = `month(visit_date, label = TRUE)`) %>%
  ggplot(data = ., aes(x = Month, y = n)) +
  geom_bar(stat = "identity")
```


```{r}
lhprism_data %>%
  group_by(species_code) %>%
  count() %>%
  arrange(desc(n))
```

What's the total amount of herbicides used?

```{r}
herb_type_tbl <- 
  lhprism_data %>% 
  filter(!is.na(herbicide_name)) %>%
  group_by(herbicide_name) %>%
  count() %>%
  arrange(desc(n)) %>%
  rename(`Herbicide Type` = herbicide_name, `Total Count` = n)
knitr::kable(herb_type_tbl, caption = "Counts for herbicide types used")

lhprism_data %>% 
  filter(!is.na(herbicide_name)) %>%
  group_by(herbicide_name, product_units) %>%
  summarise(total_herb_vol = sum(amt_of_product, na.rm = TRUE)) %>%
  arrange(desc(total_herb_vol))
```

```{r}

lhprism_data %>%
  group_by(site) %>%
  summarise(
    mean_plant_count = mean(plant_count, na.rm = TRUE),
    #mean_percent_cover = mean(percent_cover, na.rm = TRUE),
    total_visits = n()
  ) %>%
  arrange(desc(total_visits))
```

