---
title: "LHPRISM Exploratory Analysis"
format: 
  docx:
    toc: true
editor: visual
---

## Overview

This document goes through various exploratory analyses of the LHPRISM post-treatment invasive species survey data set.

## Data cleaning

This step was completed using OpenRefine. The JSON code for all steps can be found in the file `data/lhprism_openrefine_cleaning.json`. The key data cleaning steps and assumptions taken are listed below.

-   LIST KEY DATA CLEANING STEPS AND ASSUMPTIONS
-   MAKE SURE TO POINT OUT ASSUMPTIONS ASSOCIATED WITH PERCENTAGE CATEGORIES

## Reading in the data

The data are provide to us as an Excel file with multiple sheets. We are most interested in the sheet named "Visits". We will use the `readxl` package to read in the data into our work environment. But first, we will use the `pacman` package to load, and install if needed, any other packages. If this is the first time you are using `pacman`, remember that you first need to install this package.

```{r}
## Get packages
library(pacman)
pacman::p_load(here, tidyverse, janitor, lubridate)

## Get the project level directory path
proj_path <- here()

## Define a color blind friendly palette
## These are from - http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

In this code chunk, we will read in the data file. You may have to edit this call, depending on where you have saved your version of the Excel file.

```{r}
## Read in the data file
lhprism_data <- read_csv(paste0(proj_path, "/data/LHPRISM-Data.csv"), 
    col_types = cols(`iMap Treat` = col_number(), 
        GROSS_TREAT_SQFT = col_number())) %>%
  clean_names()

## Edit the order of levels for some of the variables
lhprism_data <- lhprism_data %>%
  mutate(native = factor(native, levels = c("low", "medium", "high")),
         non_nat = factor(non_nat, levels = c("low", "medium", "high")))

## Create a new species code column
lhprism_data$species_code <- str_extract(lhprism_data$site_code, "(?<=_)[^_]+(?=_)") 
## Remove plant column
lhprism_data$plant <- NULL

## Create a new site column using the species_code column
lhprism_data$site <- str_extract(lhprism_data$site_code, "^[^_]+")

## Remove all rows that have NA for site
lhprism_data <- lhprism_data %>%
  filter(!is.na(site))

# Make a column for number of visits for each Site Code.
lhprism_data <- lhprism_data %>%
  group_by(site_code) %>%
  mutate(tot_visits = length(unique(visit_date)))


```

## Figures

```{r}
#| label: fig-native-vs-nonnative-coverage
#| fig-cap: "Relationship between native vs non-native coverage categories"
#| fig-width: 6


lh_sub <- lhprism_data %>%
  select(native, non_nat)

lh_sub_long <- 
  data.frame(native_vs_nonnative = rep(c("native", "non_nat"), each = nrow(lh_sub)),
             coverage = c(lh_sub$native, lh_sub$non_nat))

lh_sub_long %>%
  filter(!is.na(coverage)) %>%
  ggplot(., aes(x = coverage, fill = native_vs_nonnative)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = cbPalette) +
  xlab("Coverage Category") +
  ylab("Visit Count") +
  labs(fill = "Coverage Type") +
  theme_minimal()
```

Let's make a table that goes along with this.

```{r}
#| label: tbl-native-vs-nonnative-coverage-table

native_vs_nonnative_cover <-
  lhprism_data %>% 
  group_by(native, non_nat) %>%
  count() %>%
  filter(!is.na(native)) %>%
  pivot_wider(names_from = non_nat, values_from = n, values_fill = 0)

knitr::kable(native_vs_nonnative_cover,
             caption = "Counts of native vs non-native coverage categories. Note that rows are native classificaitons and columns are non-native classifications.")

## Perform a chi-squared analysis
native_vs_nonnative_cover_chisq <-
  chisq.test(native_vs_nonnative_cover[2:4])

native_vs_nonnative_cover_chisq
native_vs_nonnative_cover_chisq$residuals
```

### Histogram of visit date

```{r}
ggplot(data = lhprism_data, aes(x = visit_date)) +
  geom_histogram()
```

### Histogram of pu_count

```{r}
ggplot(data = lhprism_data, aes(x = pu_count)) +
  geom_histogram() +
  scale_x_log10()
```

Let's create a new variable that is the log10 of the pull-up count

```{r}
lhprism_data <- lhprism_data %>%
  mutate(log_pu_count = log10(pu_count))

# Replace -Inf with NA
lhprism_data$log_pu_count[is.infinite(lhprism_data$log_pu_count)] <- NA

lhprism_data %>% 
  filter(pu_count > 0) %>%
  ggplot(data = ., aes(x = log_pu_count)) +
  geom_histogram()
```

Calculate the mean and confidence interval for the log_pu_count values

```{r}
log_pu_count_mean <- mean(lhprism_data$log_pu_count, na.rm = TRUE)
log_pu_count_sd <- sd(lhprism_data$log_pu_count, na.rm = TRUE)

log_pu_count_upperCI <- log_pu_count_mean + 1.96 * log_pu_count_sd / sqrt(sum(!is.na(lhprism_data$log_pu_count)))
log_pu_count_lowerCI <- log_pu_count_mean - 1.96 * log_pu_count_sd / sqrt(sum(!is.na(lhprism_data$log_pu_count)))

# Convert mean back to original scale
10^log_pu_count_mean

# Compare this to the mean and median of the original pu_count variable
mean(lhprism_data$pu_count, na.rm = TRUE)
median(lhprism_data$pu_count, na.rm = TRUE)

```

Is this different than what a group a volunteers may be able to pull-up in a day?

```{r}
volunteer_pu_count <- log10(150) # Estimate based on quick AI search
```

How many SDs away is the volunter pull-up count from our observed mean?

```{r}
(volunteer_pu_count - log_pu_count_mean) / log_pu_count_sd
```

What is the probability of observing our mean or something more extreme?

```{r}
2 * pnorm(q = abs((volunteer_pu_count - log_pu_count_mean) / log_pu_count_sd), lower.tail = FALSE)
```

### Relationship between gross square feet searched vs treated

```{r}
ggplot(data = lhprism_data, aes(x = gross_sqft, y = gross_treat_sqft)) +
  geom_point()
```

## Tables

Let's make some tables.

```{r}
#| label: tbl-visit-status-table
#| fig-cap: "Counts of visit status categories. Note that the 'NA' category is for visits where the visit status was not recorded."

lhprism_data %>%
  group_by(visit_status_char) %>%
  count() %>%
  filter(!is.na(visit_status_char)) %>%
  arrange(desc(n)) %>%
  rename(tot_count = n)

lhprism_data %>%
  group_by(visit_status_num) %>%
  count() 

lhprism_data %>%
  group_by(percent_treated) %>%
  count()

lhprism_data %>%
  group_by(distribution) %>%
  count()

lhprism_data %>%
  count(month(visit_date, label = TRUE)) %>%
  filter(!is.na(`month(visit_date, label = TRUE)`)) %>%
  rename(Month = `month(visit_date, label = TRUE)`) %>%
  ggplot(data = ., aes(x = Month, y = n)) +
  geom_bar(stat = "identity")
```

```{r}
lhprism_data %>%
  group_by(species_code) %>%
  count() %>%
  arrange(desc(n))
```

What's the total amount of herbicides used?

```{r}
herb_type_tbl <- 
  lhprism_data %>% 
  filter(!is.na(herbicide_name)) %>%
  group_by(herbicide_name) %>%
  count() %>%
  arrange(desc(n)) %>%
  rename(`Herbicide Type` = herbicide_name, `Total Count` = n)
knitr::kable(herb_type_tbl, caption = "Counts for herbicide types used")

lhprism_data %>% 
  filter(!is.na(herbicide_name)) %>%
  group_by(herbicide_name, product_units) %>%
  summarise(total_herb_vol = sum(amt_of_product, na.rm = TRUE)) %>%
  arrange(desc(total_herb_vol))
```

```{r}

lhprism_data %>%
  group_by(site) %>%
  summarise(
    mean_plant_count = mean(plant_count, na.rm = TRUE),
    #mean_percent_cover = mean(percent_cover, na.rm = TRUE),
    total_visits = n()
  ) %>%
  arrange(desc(total_visits))
```

## Change through time

First, let's find out which species were noted as eradicated. Note that eradicated is considered any site that has no target plants found upon search on the third year after treatment, which corresponds to a `visit_status_num` value of 5.

There are `r length(unique(lhprism_data$species_code))` unique species codes in the data set.

Of these, only three have any populations that were eradicated.

```{r}
unique(lhprism_data$species_code) # 85

lhprism_data %>%
  filter(visit_status_num == 5) %>%
  select(species_code) %>%
  unique()
```

Let's get some summaries.

```{r}
lhprism_data %>%
  filter(!is.na(species_code)) %>%
  group_by(species_code) %>%
  summarise(
    unique_populations = n_distinct(site_code),
    total_visits = n(),
    highest_visit_status = max(visit_status_num, na.rm = TRUE)
  ) %>%
  arrange(desc(total_visits))
```

## Practice ANOVA

```{r}
focal_spec <- c("CYSC4", "AREL8", "PUMO", "SAGL12", "HEMA17")

lhprism_data %>%
  filter(species_code %in% focal_spec) %>%
  ggplot(data = ., aes(x = species_code, y = log_pu_count)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha =0.4) +
  scale_fill_manual(values = cbPalette) +
  xlab("Species Code") +
  ylab("Log10 Pull-up Count") +
  theme_minimal()

focal_spec_lm <-
  lhprism_data %>%
  filter(species_code %in% focal_spec) %>%
  lm(log_pu_count ~ species_code, data = .) 

summary(focal_spec_lm)
anova(focal_spec_lm)
```

**Critical thinking question:** Considering the assumptions of ANOVA (normality, homogeneity of variance, independence), do you think this is an appropriate test to use for this data set? Why or why not? If not, what other test(s) might be more appropriate?

## Practice linear regression

```{r}
# examine relationship between visit_status_num and gross_sqft
ggplot(data = lhprism_data, aes(x = gross_sqft, y = gross_treat_sqft)) +
  geom_point() +
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method = "lm", se = FALSE, color = "red")

# fit a linear regression model
lm_model <- lm( log10(gross_treat_sqft + 0.1) ~ log10(gross_sqft +0.1), data = lhprism_data)
summary(lm_model)
```

```{r}
# example of a situation where regression *could* be used, but quantile regression might be more appropriate
ggplot(data = lhprism_data, aes(x = gross_sqft, y = log_pu_count)) +
  geom_point() +
  scale_x_log10()
```
